Categorical Predictor Embeddings in Actuarial Modeling

### 2021-03-11 manuscript/manuscript.Rmd
models

---
title: Categorical Predictor Embeddings in Actuarial Modeling
authors:
  - name: Kevin Kuo 
    affiliation: RStudio and Kasa AI
    email: kevin@kasa.ai
  - name: Ronald Richman
    affiliation: QED Actuaries and Consultants
    email: ronaldrichman@gmail.com
abstract: |
  Enter the text of your abstract here.
keywords:
  - blah
  - blee
  - bloo
  - these are optional and can be removed
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
header-includes: |
    \usepackage{amsmath}
    \usepackage{mathtools}
    \usepackage{bm}
    \DeclarePairedDelimiter\ceil{\lceil}{\rceil}
---

@@

# Definitions {#defns}

@@

We take the log of the continuous predictor `building_insurance_coverage`, which allows the scale of the predictor to 
match that of the target variable. Because the `flood_zone` variable in the original data contains 60 levels, we take 
the prefix of the zone code, which corresponds to the level of risk as determined by FEMA [^8]. For example, `"A01"`, 
`"A02"`, and so on are recoded as simply `"A"`. A log link together with the gamma distribution is a standard choice 
for severity modeling, which provides a multiplicative structure where the response is positive.

@@

```{r nn-1, echo = FALSE, fig.cap = "Architecture of Models 2 and 4.", out.width = "90%", fig.align = "center"}
knitr::include_graphics("images/nn.png")
```

@@

For Model 3, we return to the GLM, but now we replace the categorical variables with the trained embeddings from Model
2. In other words, our model now contains only continuous variables, and each categorical factor is represented as a
scalar value and obtains its own coefficient. (TODO: insert formal eqn for mean)

@@

Model 4 is architecturally identical to Model 2, with the only difference being the number of embedding dimensions:
instead of mapping factor levels to real numbers, we map them to a point in Euclidean space, where the dimension of the space is the ceiling
of the factor's cardinality divided by two. Formally, in the notation of Section \ref{catdata},
$q_{\mathcal{P}^j}=\ceil{n^j_{\mathcal{P}}/2}$ for variable $X_j$.

>>>>

---
title: Categorical Predictor Embeddings in Actuarial Modeling
authors:
  - name: Kevin Kuo 
    affiliation: RStudio and Kasa AI
    email: kevin@kasa.ai
  - name: Ronald Richman
    affiliation: QED Actuaries and Consultants
    email: ronaldrichman@gmail.com
abstract: |
  Enter the text of your abstract here.
keywords:
  - blah
  - blee
  - bloo
  - these are optional and can be removed
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
header-includes: |
    \usepackage{amsmath}
    \usepackage{mathtools}
    \usepackage{bm}
    \usepackage{bbm}
    \DeclarePairedDelimiter\ceil{\lceil}{\rceil}
---

@@

# Theoretical Overview {#defns}

@@

We take the log of the continuous predictor `building_insurance_coverage`, which allows the scale of the predictor to 
match that of the target variable. Because the `flood_zone` variable in the original data contains 60 levels, we take 
the prefix of the zone code, which corresponds to the level of risk as determined by FEMA [^8]. For example, `"A01"`, 
`"A02"`, and so on are recoded as simply `"A"`. A log link together with the gamma distribution is a standard choice 
for severity modeling, which provides a multiplicative structure where the response is positive. Following the notation
in \@ref(catdata), the number of parameters in the GLM is

\begin{equation}
1 + |J^{\text{num}}| + \left(\sum_{j \in J^{\text{cat}}} n_{\mathcal{P}}^j\right) - |J^{\text{cat}}|:
(\#eq:glm1)
\end{equation}

one for the intercept, one for each numeric variable, and the numbers of levels minus one (due to dummy coding) for 
each categorical variable.

@@

```{r nn-1, echo = FALSE, fig.cap = "Architecture of Models 2 and 4.", out.width = "80%", fig.align = "center"}
knitr::include_graphics("images/nn.png")
```

@@

For Model 3, we return to the GLM, but now we replace the categorical variables with the trained embeddings from Model
2. In other words, our model now contains only continuous variables, and each categorical factor is represented as a
scalar value and obtains its own coefficient. In contrast with the number of parameters computed in Equation
\@ref(eq:glm1), we have fewer parameters for this model:

\begin{equation}
1 + |J^{\text{num}}| + |J^{\text{cat}}|.
(\#eq:glm2)
\end{equation}

Hence, the model can be written as

\begin{equation}
\mu_i = \beta_0 + \sum_{j \in J^{\text{num}}}\beta_j x_{i,j} + \sum_{j \in J^{\text{cat}}} \beta_j z_{i,j},
\end{equation}

where $z_{i,j}$ represents the embedding value for $i$th instance of the categorical column $X_j$.

Although in this particular example we incorporate one-dimensional embeddings, one can easily include multidimensional
embeddings, which would increase the model's flexibility. The modeler would need to balance the additional lift with
some loss of interpretability, since with one-dimensional embeddings, our GLM structure dictates that there is a 
monotonic relationship between the predicted response and the input embedding, which could be desirable in certain
use cases (and undesirable in others.)

Since we only have a few parameters in this model, we exhibit the full list of fitted coefficients in Table 
\@ref(tab:glm2).

Table: (\#tab:glm2) Coefficients of Model 3.

|term                                     | estimate|
|:----------------------------------------|--------:|
|(Intercept)                              |    2.389|
|total_building_insurance_coverage        |    0.568|
|basement_enclosure_crawlspace_type       |    0.136|
|number_of_floors_in_the_insured_building |   -0.024|
|occupancy_type                           |    0.067|
|flood_zone                               |    0.035|
|primary_residence                        |   -0.001|
|community_rating_system_discount         |    0.504|

@@

Model 4 is architecturally identical to Model 2, with the only difference being the number of embedding dimensions:
instead of mapping factor levels to real numbers, we map them to points in Euclidean space, where the dimension of the 
space is the ceiling
of the factor's cardinality divided by two. Formally, in the notation of Section \ref{catdata},
$q_{\mathcal{P}^j}=\ceil{n^j_{\mathcal{P}}/2}$ for variable $X_j$. Following the practice from the previous three
models, we list the fitted embeddings for a couple variables in Tables \@ref(tab:nn2-occupancy-type) and 
\@ref(tab:nn2-flood-zone). Note that, for `flood_zone`, we are
only displaying the embedding for *one* level.

Table: (\#tab:nn2-occupancy-type) Learned embeddings for `occupancy_type` in Model 4.

|level                                       |    e1|    e2|   e3|
|:-------------------------------------------|-----:|-----:|----:|
|2 to 4 unit residential buidling            | -1.44| -1.97| 2.35|
|Non-residential building                    | -6.83| -8.11| 5.34|
|Residential building with more than 4 units | -7.63| -6.58| 5.09|
|Single family residence                     | -1.61| -2.54| 3.01|

Table: (\#tab:nn2-flood-zone) Learned embeddings for level `"A00"` of variable `flood_zone` in Model 4.

|dimension |value  |dimension |value  |dimension |value  |dimension |value  |
|:---------|:------|:---------|:------|:---------|:------|:---------|:------|
|e1        |0.020  |e11       |2.146  |e21       |-1.933 |e31       |-0.941 |
|e2        |-0.885 |e12       |1.723  |e22       |1.150  |e32       |-1.279 |
|e3        |-0.126 |e13       |0.378  |e23       |0.863  |e33       |-0.360 |
|e4        |1.491  |e14       |-0.310 |e24       |-0.471 |e34       |-1.173 |
|e5        |0.980  |e15       |2.482  |e25       |0.929  |          |       |
|e6        |0.952  |e16       |1.823  |e26       |-1.139 |          |       |
|e7        |-0.217 |e17       |-2.715 |e27       |0.113  |          |       |
|e8        |-0.611 |e18       |1.144  |e28       |1.155  |          |       |
|e9        |0.919  |e19       |-0.045 |e29       |0.707  |          |       |
|e10       |-1.196 |e20       |0.467  |e30       |1.358  |          |       |
